Chasing Speeders: Analyzing Toronto’s 2023 Speeding Trends in School Zones”

### 2024-12-07 paper/paper.qmd
improve/fix wording, formatting, extra text

---
title: "Chasing Speeders: Analyzing Toronto’s 2023 Speeding Trends in School Zones”"
subtitle: "Volume Versus Speed Limits: 0.12 km/h Declines That Add Up Across School-Traffic Zones"
author: 
  - Karen Riani
thanks: "Code and data are available at: https://github.com/karenrni/Analyzing-Torontos-Speeders/."
date: today
date-format: long
abstract: "This study investigates over-speeding in Toronto, defined as the amount by which vehicles exceed posted speed limits, using data from the city's School Safety Zone Watch Your Speed Program and Traffic Camera datasets. A multi-linear regression analysis revealed key drivers of speeding behavior: higher traffic volumes reduce over-speeding by 0.12 km/h per additional vehicle, while increasing speed limits correlate with a 0.13 km/h rise in excess speed per km/h limit increase. Geographic variability was significant, with Ward 4 showing a 4.8 km/h reduction in over-speeding compared to the reference ward. Despite these findings, the negligible impact of speed cameras highlights limitations in existing enforcement data. These insights underscore the importance of nuanced, data-driven approaches to traffic management and policy development in urban settings."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
fig_caption: yes
---

@@

The rise of advanced vehicle technology and the adrenaline appeal of street racing exacerbate these issues, with young male drivers (18–25 years) disproportionately represented in this trend. Though rare at **0.9% prevalence**, street racing poses a **five-fold increase in crash risk (OR = 5.23, p < .001)**, even after accounting for other risk factors [@simonsmorton2017]. Despite Ontario’s strict penalties, such as **$10,000 fines** and **30-day license suspensions for excessive speeding**, enforcement alone has not fully addressed these challenges [@trafficparalegal2024]. Recent headlines, such as a driver clocked at **102 km/h in a 50 km/h school zone** and a Toronto speed camera being destroyed less than 24 hours after reinstallation, underscore the disconnect between enforcement measures and public behavior [@citynews2024; @cbc2023].

This paper focuses on understanding the estimand **over_speed_limit**, the amount by which speeds in each observed speed bin exceed the posted speed limit in 2023. This dependent variable is calculated as the difference between the lower bound of the observed speed bin and the speed limit, ensuring a straightforward measure of speeding severity. By quantifying how far drivers exceed the limit, over_speed_limit allows for identifying patterns of high-risk speeding behaviour. Using open data from Toronto's School Safety Zone Watch Your Speed Program, Watch Your Speed Detailed Counts, and Traffic Camera datasets, including vehicle counts (volume) in each speed bin, speed limits, and ward-level identifiers, this study investigates how traffic patterns, enforcement mechanisms, and geographic factors contribute to variations in speeding behaviour across the city.

To address this, the study employs multi-linear regression models to examine the relationship between excess speed and predictors such as traffic volume, speed limits, and ward characteristics throughout 2023. Results indicate that higher traffic volume reduces excess speed by **0.12 km/h per additional vehicle**, while speed limits positively influence speeding by **0.13 km/h per unit increase**. Wards exhibit substantial variability, with some, such as Ward 4, showing significant decreases in speeding. However, the impact of individual speed cameras was negligible, suggesting limitations in the data and methodology for evaluating camera-specific effects.

@@

## Overview
The data for this analysis is sourced from Open Data Toronto [@opendatatoronto] and comprises three datasets: School Safety Zone Watch Your Speed Program – Locations [@schoolzonelocations], School Safety Zone Watch Your Speed Program – Detailed Speed Counts in 2023 [@schoolzonespeeds], and the Traffic Camera dataset  [@trafficcameras]. These datasets provide detailed insights into speeding behavior, traffic enforcement mechanisms, and geographic characteristics within Toronto. The traffic camera and speed sign data were merged through a spacial join with a range of 500m using geographical coordinates. This was joined with the detailed counts of speeds through Toronto ward number (25 ward structure).

## Packages Used
The analysis utilized R [@citeR] with the following packages: `opendatatoronto` [@opendatatoronto], `tidyverse` [@tidyverse], `janitor` [@janitor], `arrow` [@arrow], `readr` [@readr], `sf` [@sfpackage], `testthat` [@testthat], `ggplot2` [@ggplot2_wickham_2016], `corrplot` [@corrplot], `tidymodels` [@tidymodels], `modelsummary` [@modelsummary], `kableExtra` [@kableExtra] and `rstanarm` [@rstanarm].

@@

The datasets from Open Data Toronto serve as the foundation for analyzing speeding behavior in Toronto. Each variable is derived from real-world phenomena and represents a specific aspect of traffic patterns or enforcement mechanisms:

- **Over Speed Limit**: This dependent variable quantifies the extent to which drivers exceed posted speed limits. It is derived by calculating the difference between the midpoint of the observed speed bin (e.g., `[45,50)` translates to a midpoint of 47.5 km/h) and the posted speed limit. This method provides a measure of speeding severity, distinguishing between minor infractions and extreme violations.

- **Sign ID and Camera ID (X_id)**: These unique identifiers represent speed cameras and their locations, facilitating the mapping of speeding patterns to enforcement points. However, sign addresses may not always align precisely with camera placements, as discussed in the limitations section.

- **Longitude and Latitude**: Geographic coordinates capturing the locations of observed speeding behaviour, allowing for spatial analysis and the identification of high-risk zones.

- **Speed Limit**: The posted speed limits for each location, ranging from 30 km/h to 50 km/h, with an average of 38.24 km/h. This variable provides the baseline against which speeding is measured.

- **Volume**: The count of vehicles observed in a given speed bin, representing traffic density. Median volume is 13 vehicles, with counts ranging widely, up to a maximum of 209 vehicles.

- **Speed Bin**: Recorded ranges of observed speeds, such as `[45,50)`, provide context for speed categories, assuming uniform speed distributions within each bin.

- **Ward Number**: Geographic zones represented as categorical identifiers, reflecting administrative boundaries within Toronto. For instance, Ward 10 accounts for 28.8% of observations, making it the most represented ward, while Ward 19 has considerably fewer observations.

- **No Camera in Radius**: This binary variable was constructed based on spatial joins between speed observation data and camera locations. If a given location fell outside the defined radius of nearby cameras, it was flagged as `TRUE`. This captures the absence of enforcement presence and allows for an analysis of its influence on speeding behavior.

- **Camera Density**: To represent enforcement intensity across wards, the number of cameras was aggregated per ward. This variable offers insight into how concentrated enforcement might correlate with traffic patterns or speeding violations.

@@

The datasets used in this study have notable limitations that must be acknowledged. **Sign addresses** in the datasets have not been verified and may represent the address of an adjacent property rather than the actual location of the sign. The **speed limit** column reflects the most recent operating parameter, which does not necessarily align with the current or previous speed limits for each location.

The **Traffic Cameras** dataset is highly up-to-date, with a last refresh on 2024-11-26. While there are no issues with accessibility or completeness, usability is constrained by unclear column names and incomplete metadata. Conversely, the **School Safety Zone Watch Your Speed Program – Detailed Speed Counts** dataset is over 12 months outdated, requiring reliance on direct CSV links for access. This reduces its applicability for analysing current traffic trends.

To address gaps, **spatial joins** were employed to map camera locations to speed observation points, identifying enforcement presence, and **calculated variables**, such as over speed limit and camera density, were derived to facilitate analysis. However, these transformations have limitations. Speed bin ranges assume a uniform speed distribution, which may not capture the variability in driver behaviours. Furthermore, camera placements are concentrated in high-traffic or high-risk areas, potentially underrepresenting quieter zones. Together, these factors highlight the need for caution when interpreting results, as the datasets provide a snapshot rather than a comprehensive depiction of traffic behaviour.

## Descriptive Statistics
Descriptive statistics in [@tbl-summary] highlight critical patterns within the data:

- The **mean over speed limit** is 3.91 km/h, with extreme violations reaching up to 70 km/h.

- The **median speed bin midpoint** is 30 km/h, while the mean traffic volume is 23.35 vehicles.

- Geographic variables such as **longitude** and **latitude** center around Toronto’s urban core.

@@

```
[fig-heatbar]  visualizes the total volume of over-speeding incidents across various wards. Ward 10 stands out prominently with the highest number of over-speeding incidents, exceeding 1,500, which may indicate unique characteristics such as higher traffic density or insufficient enforcement measures. Other wards, such as Wards 13 and 16, also show elevated levels of over-speeding but fall significantly behind Ward 10.

@@

```
The scatter plot in [fig-cam-dens] explores the relationship between the number of speed cameras nearby and the average speed over the limit across various wards. The slight downward trend, indicated by the fitted regression line, suggests a weak negative association between the number of cameras and over-speeding behavior. Specifically, as the number of cameras increases, the average speed over the limit tends to decrease, albeit marginally. However, the variability in average speed over the limit across wards with the same number of cameras highlights the potential influence of other ward-specific factors, such as enforcement policies or road design.

@@

[@fig-heatmap1] provides a detailed visualization of speeding behavior across Toronto’s wards, showcasing the relationship between the extent of over-speeding (measured in km/h above the limit) and the total volume of vehicles recorded. Ward 10 stands out with the highest total speeding volume, concentrated in lower over-speeding categories (<40 km/h above the limit), indicating significant non-compliance with posted limits. Conversely, instances of extreme over-speeding (>50 km/h above the limit) are sparse but occur across wards 4, 13, and 25, highlighting potential hotspots for dangerous driving behavior.

@@

This section outlines the modeling process used to analyze over-speeding behavior based on data from Toronto’s School Safety Zone Watch Your Speed Program and Traffic Cameras datasets. The goal was to predict the extent to which observed speeds exceeded posted limits (over_speed_limit) by evaluating various predictors. Three models were developed and assessed using performance metrics, including RMSE, MAE, and \( R^2 \), to determine the most appropriate model. The final model selection was based on its balance between simplicity, accuracy, and interpretability.

@@

- Model 1: Includes ward-level categorical effects (ward_no) to account for geographic differences in speeding behavior.

@@

The selection process aimed to balance complexity and explanatory power. Ward-level effects were included to capture geographic variations, while Camera Identifiers were tested to evaluate their influence. Redundant variables were excluded to maintain simplicity. Model 1 was selected as the final model because it demonstrated superior performance across validation metrics. By including ward-level effects, it effectively captured geographic differences in speeding behaviour, making it the most suitable choice.

@@

The RMSE for Model 1 was 7.01, indicating that, on average, predictions deviated from observed values by this amount. MAE was 4.77, highlighting its precision in predicting over-speeding behavior. The \( R^2 \) value of 0.176 showed that the model explained 17.6% of the variance in over-speeding behavior, providing a reasonable level of explanatory power for the observed data.

@@

Variance Inflation Factors (VIFs) were calculated to evaluate multicollinearity, with all variables in Model 1 displaying acceptable values below 4. Although Model 3 included Camera Identifiers, this variable was statistically insignificant and did not enhance model performance. Model 1 effectively balanced complexity and accuracy, integrating ward-level effects to provide valuable insights into over-speeding behavior while maintaining interpretability. Consequently, Model 1 was selected as the final model.

@@

The \(R^2\) value of 0.176 and adjusted \(R^2\) of 0.175 show that the model explains approximately 17.6% of the variance in over-speeding. Although relatively modest, this reflects the challenge of modeling complex urban traffic behaviors with diverse contributing factors.

#### Statistical Fit Metrics
With an AIC of 99,722.4 and a BIC of 99,851.4, the model, these values are consistent with the dataset's size, being 14,594 observations.The model's F-statistic of 207.459 shows the statistical significance of the predictors in explaining variations in over-speeding. This shows that the inclusion of speed limits, traffic volume, and ward-level effects contributes significantly to the model’s explanatory power.

@@

```{r}
#| label: fig-observed-vs-predicted
#| fig-cap: "Observed vs. Predicted Over-Speeding: Model Accuracy Visualization"

@@

[@fig-observed-vs-predicted] shows the relationship between predicted and observed over-speeding values. Predicted values align reasonably well with observed values at lower ranges (0–10 km/h). This indicates that the model performs adequately in capturing moderate over-speeding behaviors, where most of the data points fall close to the red dashed line representing perfect predictions.

@@

While the model provides reasonable predictions for average over-speeding behaviors, its performance diminishes at higher observed values, indicating potential limitations in its ability to capture outliers or extreme behaviors.

@@

### Wards with Lower Predictions
In contrast, Wards 5, 21, and 20 show substantially lower predicted over-speeding values, averaging below 3 km/h. These results suggest that predicted over-speeding behaviors are less prominent in these wards, potentially influenced by traffic conditions, enforcement presence, or other unobserved factors.

#### Implications of Predictions
While the predictions align with observed patterns in some wards, such as higher over-speeding in Ward 12, the variability across wards suggests that spatial characteristics may play a significant role. These predictions provide a baseline understanding of over-speeding behavior at a ward level.

@@

## Ward-Level Disparities in Over-Speeding
The results show geographical disparities in over-speeding across Toronto. For instance, Ward 12 emerges as a statistical outlier, with an average predicted over-speeding of 12.5 km/h. In  contrast, Ward 5 shows much lower predictions, averaging less than 3 km/h. This stark difference raises questions about factors such as road design, traffic enforcement, and even demographic variations. Ward 12 may be plagued by wide arterial roads inviting speed or does has outdated enforcement options. The low over-speeding in Ward 5 may hint at effective urban planning or stronger enforcement presence. It’s clear that “where” you drive in Toronto heavily influences “how” you drive. However, wards span cities in the 25 Ward model, and generalizing spaces of this size may also look over more specific area data.

Ward 4, which includes Parkside Drive—Toronto's most-ticketed location for speeding [@cbc2023], exhibited an intriguing pattern. Despite its reputation for dangerous speeding and expected dominance in over all speeding, the model indicated that over-speeding in Ward 4 was, on average, 4.77 km/h lower than the reference Ward 2. This discrepancy suggests the effectiveness of enforcement measures or specific traffic conditions in mitigating speeding behavior.

@@

## Volume of Speeds
Surprisingly, higher traffic volumes seem to correlate with lower over-speeding rates. The consistently negative coefficient for volume across all models suggests that when more vehicles are on the road at any speed, the likelihood of speeding diminishes. One possible explanation is that congestion creates natural constraints, limiting a driver’s ability to exceed the speed limit alongside the speed of their drivers around them.

## Outliers 
The scatterplot of predicted vs. observed over-speeding values in [@fig-ward-predictions] reveals areas where the model struggles, particularly with extreme outliers. Predictions for high over-speeding values (e.g., above 40 km/h) are generally underrepresented, while predictions cluster heavily near lower values. This suggests the model may be oversimplifying the dynamics of extreme over-speeding events, which likely involve additional, unmeasured factors like driver behavior or road conditions. Additionally, the residual errors in [@fig-residuals] highlight slight over-predictions for certain wards, hinting at potential biases in how ward-level effects were estimated.

@@

### Data and Method Limitations
The analysis does not account for date and time-specific patterns. Over-speeding frequently occurs during night hours or non-peak times when roads are less congested and school zone speeds only ooperate between 8 AM and 5 PM. Additionally, it is unclear whether the speed signs are specifically tracking school zone speeds or normal speeds. Some speed cameras included in the dataset were taken down partway through 2023, but they were retained in the analysis as long as they existed at any point during the year. While this decision did not significantly affect results, it did not clearly reflect temporal changes in camera presence.

The School Safety Zone Watch Your Speed Program sensors may occasionally track outliers such as ambulances or police vehicles, leading to exaggerated over-speeding measurements. Additionally, these systems do not log vehicle types, which could have provided more support into driving patterns. The analysis assumes that vehicles maintain a constant speed through the sensor area, which may not reflect real-world driving behaviour. Traffic patterns can be complex and consequently difficult to accurately model. The multi-linear regression model used could benefit from more layered or bayesian models and uncertainty handling.

A major assumption in the modeling process is the uniform distribution of speeds within speed bins. However, this may not reflect real-world conditions, where driver behavior often skews towards certain thresholds (e.g., staying just below enforcement limits). Similarly, this analysis aimed to examine trends in speeding around schools near traffic cameras, however it lacks in tracking speed demons and focuses on all speeds above the limit.

@@

## Next Steps and recommendations

Future models could include weather data, as adverse conditions (e.g., rain, snow) are known to impact driving behavior. Open Data Toronto provides weather datasets that could be merged with speeding metrics for more comprehensive analyses, if joins are possible. Furthermore, considering the time of speed logging may improve and further detail the analysis, as speeding increases towards the peak hours and the night time [@nightSpeed] and it would better align with Toronto school zone speed laws from 8am-5pm.

@@

Incorporating data on vehicle types could help identify whether certain vehicles (e.g., trucks, commercial vehicles) contribute disproportionately to over-speeding. Modified cars tend to speed at greater velocity and more often than others [@], offering additional insights into driver behaviour. Furthermore, there has been evidence that increases in vehicle height also correlate with speeding, as vehicle height may affect drivers' speed perception [@height].

As speeding trends can vary greatly over time, addtional analysis spanning over several years can further support driver trends and behaviour. For example, speeding increased during the pandemic, likely due to the significant decrease in cars on the road, offering more leeway for speeding [@pandemic]. Having a stronger focus on speed demons may be of interest, as tracking driver behaviour of those who minimally speed may not be as telling of general speeder habits.

@@

## Overview
This appendix outlines a survey methodology designed to improve the analysis of excessive speeding near Toronto school zones. With a $500K budget, this approach aims to gather data from drivers, pedestrians, and community members about speeding behaviours, awareness of enforcement measures, and perceived safety. By joining observational data with survey insights, this methodology aims to improve analysis and informed policy recommendations.

@@

Vision Zero’s annual budget for public engagement and data-driven initiatives often ranges between 1-2 million, justifying a scaled-down 500K budget for this specific study.Based on this similar city-wide traffic safety initiatives the $500K is allocated to account for:

@@

- Advanced recruitment and incentivization strategies to ensure adequate participation.

@@

**Drivers:**  
- Demographic: Ages 18-40, with emphasis on young male drivers.

- Behavioral patterns: Frequent commuters in school zones.

**Pedestrians:**  
- Parents, school staff, and local residents.

**Community Advocates:**  
- Stakeholders involved in traffic safety initiatives.

>>>>

---
title: "Chasing Speeders: Analyzing Toronto’s 2023 Speeding Trends in School Zones”"
subtitle: "Volume Versus Speed Limits: 0.12 km/h Declines That Add Up Across School-Traffic Zones"
author: 
  - Karen Riani
thanks: "Code and data are available at: https://github.com/karenrni/Analyzing-Torontos-Speeders/."
date: today
date-format: long
abstract: "This study investigates over-speeding in Toronto, defined as the amount by which vehicles exceed posted speed limits, using data from the city's School Safety Zone Watch Your Speed Program and Traffic Camera datasets. A multi-linear regression analysis revealed drivers of speeding behaviour near schools: higher traffic volumes reduce over-speeding by 0.12 km/h per additional vehicle, while increasing speed limits correlate with a 0.13 km/h rise in excess speed per km/h limit increase. Geographic variability was significant, with Ward 4 showing a 4.8 km/h reduction in over-speeding compared to the reference ward. Despite these findings, the negligible impact of speed cameras highlights limitations in existing enforcement data. It highlights the need for better data to address traffic challenges effectively. While the results offer some direction, more comprehensive datasets could lead to stronger and more actionable policy recommendations."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
fig_caption: yes
---

@@

The rise of advanced vehicle technology and the adrenaline appeal of street racing exacerbate these issues, with young male drivers (18–25 years) disproportionately represented in this trend. Though rare at 0.9% prevalence, street racing poses a five-fold increase in crash risk (OR = 5.23, p < .001), even after accounting for other risk factors [@simonsmorton2017]. Despite Ontario’s strict penalties, such as $10,000 fines and 30-day license suspensions for excessive speeding, enforcement alone has not fully addressed these challenges [@trafficparalegal2024]. Recent headlines, such as a driver clocked at **102 km/h in a 50 km/h school zone and a Toronto speed camera being destroyed less than 24 hours after reinstallation, showing the disconnect between enforcement measures and public behaviour [@citynews2024; @cbc2023].

This paper focuses on understanding the estimand over_speed_limit, the amount by which speeds in each observed speed bin exceed the posted speed limit around schools in 2023. This dependent variable is calculated as the difference between the lower bound of the observed speed bin and the speed limit, ensuring a straightforward measure of speeding severity. By quantifying how far drivers exceed the limit, over_speed_limit allows for identifying patterns of high-risk speeding behaviour. Using open data from Toronto's School Safety Zone Watch Your Speed Program, Watch Your Speed Detailed Counts, and Traffic Camera datasets, including vehicle counts (volume) in each speed bin, speed limits, and ward-level identifiers, this study investigates how traffic patterns, enforcement mechanisms, and geographic factors contribute to variations in speeding behaviour across the city.

To address this, the study employs multi-linear regression models to examine the relationship between excess speed and predictors such as traffic volume, speed limits, and ward characteristics throughout 2023. Results indicate that higher traffic volume reduces excess speed by 0.12 km/h per additional vehicle, while speed limits positively influence speeding by 0.13 km/h per unit increase. Wards exhibit substantial variability, with some, such as Ward 4, showing significant decreases in speeding. However, the impact of individual speed cameras was negligible, suggesting limitations in the data and methodology for evaluating camera-specific effects.

@@

## Overview
The data for this analysis is sourced from Open Data Toronto [@opendatatoronto] and comprises three datasets: School Safety Zone Watch Your Speed Program – Locations [@schoolzonelocations], School Safety Zone Watch Your Speed Program – Detailed Speed Counts in 2023 [@schoolzonespeeds], and the Traffic Camera dataset  [@trafficcameras]. These datasets provide direction into speeding behaviour near schools, traffic enforcement mechanisms, and geographic characteristics within Toronto. The traffic camera and speed sign data were merged through a spatial join with a range of 500m using geographical coordinates. This was joined with the detailed counts of speeds through Toronto ward number (25 ward structure).

## Packages Used
The analysis utilized R [@citeR] with the following packages: `opendatatoronto` [@opendatatoronto], `tidyverse` [@tidyverse], `janitor` [@janitor], `arrow` [@arrow], `readr` [@readr], `sf` [@sfpackage], `testthat` [@testthat], `ggplot2` [@ggplot2_wickham_2016], `corrplot` [@corrplot], `tidymodels` [@tidymodels], `modelsummary` [@modelsummary], `kableExtra` [@kableExtra], `knitr` [@knitr], `here` [@here], and `rstanarm` [@rstanarm].

@@

- **over_speed_limit**: This dependent variable quantifies the extent to which drivers exceed posted speed limits. It is derived by calculating the difference between the midpoint of the observed speed bin (e.g., `[45,50)` translates to a midpoint of 47.5 km/h) and the posted speed limit. This method provides a measure of speeding severity, distinguishing between minor infractions and extreme violations.

- **sign_id and X_id (Camera id)**: These unique identifiers represent speed cameras and their locations near schools, However, sign addresses may not always align precisely with camera placements, as discussed in the limitations section.

- **longitude and latitude**: Geographic coordinates capturing the locations of observed speeding behaviour, allowing for spatial analysis and the identification of high-risk zones.

- **speed_limit**: The posted speed limits for each location, ranging from 30 km/h to 50 km/h, with an average of 38.24 km/h. This variable provides the baseline against which speeding is measured.

- **volume**: The count of vehicles observed in a given speed bin, representing traffic density. The median volume is 13 vehicles, with counts ranging widely, up to a maximum of 209 vehicles.

- **speed_bin**: Recorded ranges of observed speeds, such as `[45,50)`, provide context for speed categories, assuming uniform speed distributions within each bin.

- **ward_no**: Geographic wards represented as 25 categorical identifiers, reflecting administrative boundaries within Toronto. For instance, Ward 10 accounts for 28.8% of observations, making it the most represented ward, while Ward 19 has considerably fewer observations.

- **no_camera_in_radius**: This binary variable was constructed based on spatial joins between speed observation data and camera locations. If a given location fell outside the defined radius of nearby cameras, it was flagged as `TRUE`. This captures the absence of enforcement presence and allows for an analysis of its influence on speeding behaviour. This was exclusively used during simulation.

- **camera_density**: To represent enforcement intensity across wards, the number of cameras was aggregated per ward. This variable attempts to show how concentrated enforcement might correlate with traffic patterns or speeding violations.

@@

The datasets used in this study have notable limitations that must be acknowledged. Sign addresses in the datasets have not been verified and may represent the address of an adjacent property rather than the actual location of the sign. The speed limit column reflects the most recent operating parameter, which does not necessarily align with the current or previous speed limits for each location.

The Traffic Cameras dataset is highly up-to-date, with a last refresh on 2024-11-26. While there are no issues with accessibility or completeness, usability is constrained by unclear column names and incomplete metadata. Conversely, the School Safety Zone Watch Your Speed Program – Detailed Speed Counts dataset is over 12 months outdated, requiring reliance on direct CSV links for access. This reduces its applicability for analyzing current traffic trends.

To address gaps, spatial joins were employed to map camera locations to speed observation points, identifying enforcement presence, and calculated variables, such as over speed limit and camera density, were derived to facilitate analysis. However, these transformations have limitations. Speed bin ranges assume a uniform speed distribution, which may not capture the variability in driver behaviours. Furthermore, camera placements are concentrated in high-traffic or high-risk areas, potentially underrepresenting quieter zones. Together, these factors highlight the need for caution when interpreting results, as the datasets provide a snapshot rather than a comprehensive depiction of traffic behaviour.

## Descriptive Statistics
The descriptive statistics (see [@tbl-summary]) reveal important patterns within the dataset. The average over-speeding value is 3.91 km/h, with extreme cases reaching as high as 70 km/h, illustrating the range of speeding behavior observed. The median speed bin midpoint is 30 km/h, reflecting typical speed categories captured, while the mean traffic volume stands at 23.35 vehicles, indicating the density of traffic during observations. Geographic variables, including longitude and latitude, align closely with Toronto's urban core, emphasizing the city-centric focus of the analysis.

@@

```
[@fig-heatbar]  visualizes the total volume of over-speeding incidents across various wards. Ward 10 stands out prominently with the highest number of over-speeding incidents, exceeding 1,500, which may indicate unique characteristics such as higher traffic density or insufficient enforcement measures. Other wards, such as Wards 13 and 16, also show elevated levels of over-speeding but fall significantly behind Ward 10.

@@

```
The scatter plot in [@fig-cam-dens] explores the relationship between the number of speed cameras nearby and the average speed over the limit across various wards. The slight downward trend, indicated by the fitted regression line, suggests a weak negative association between the number of cameras and over-speeding behaviour. Specifically, as the number of cameras increases, the average speed over the limit tends to decrease, albeit marginally. However, the variability in average speed over the limit across wards with the same number of cameras highlights the potential influence of other ward-specific factors, such as enforcement policies or road design.

@@

[@fig-heatmap1] provides a detailed visualization of speeding behaviour across Toronto’s wards, showcasing the relationship between the extent of over-speeding (measured in km/h above the limit) and the total volume of vehicles recorded. Ward 10 stands out with the highest total speeding volume, concentrated in lower over-speeding categories (<40 km/h above the limit), indicating significant non-compliance with posted limits. Conversely, instances of extreme over-speeding (>50 km/h above the limit) are sparse but occur across wards 4, 13, and 25, highlighting potential hotspots for dangerous driving behaviour.

@@

This section outlines the modelling process used to analyze over-speeding behaviour based on data from Toronto’s School Safety Zone Watch Your Speed Program and Traffic Cameras datasets. The goal was to predict the extent to which observed speeds exceeded posted limits (over_speed_limit) by evaluating various predictors. Three models were developed and assessed using performance metrics, including RMSE, MAE, and \( R^2 \), to determine the most appropriate model. The final model selection was based on its balance between simplicity, accuracy, and interpretability.

@@

- Model 1: Includes ward-level categorical effects (ward_no) to account for geographic differences in speeding behaviour.

@@

The selection process aimed to balance complexity and explanatory power. Ward-level effects were included to capture geographic variations, while camera Identifiers were tested to evaluate their influence. Redundant variables were excluded to maintain simplicity. Model 1 was selected as the final model because it demonstrated superior performance across validation metrics. By including ward-level effects, it effectively captured geographic differences in speeding behaviour, making it the most suitable choice.

@@

The RMSE for Model 1 was 7.01, indicating that, on average, predictions deviated from observed values by this amount. MAE was 4.77, highlighting its precision in predicting over-speeding behaviour. The \( R^2 \) value of 0.176 showed that the model explained 17.6% of the variance in over-speeding behaviour, providing a reasonable level of explanatory power for the observed data.

@@

Variance Inflation Factors (VIFs) were calculated to evaluate multicollinearity, with all variables in Model 1 displaying acceptable values below 4. Although Model 3 included camera Identifiers, this variable was statistically insignificant and did not enhance model performance. Model 1 effectively balanced complexity and accuracy, integrating ward-level effects to show over-speeding behaviour while maintaining interpretability. Consequently, Model 1 was selected as the final model.

@@

The \(R^2\) value of 0.176 and adjusted \(R^2\) of 0.175 shows that the model explains approximately 17.6% of the variance in over-speeding. Although relatively modest, this reflects the challenge of modelling complex urban traffic behaviours with diverse contributing factors.

#### Statistical Fit Metrics
With an AIC of 99,722.4 and a BIC of 99,851.4, the model, values are consistent with the dataset's size, being 14,594 observations. The model's F-statistic of 207.459 shows the statistical significance of the predictors in explaining variations in over-speeding. This shows that the inclusion of speed limits, traffic volume, and ward-level effects contributes significantly to the model’s explanatory power.

@@

```{r}
#| label: fig-observed-vs-predicted
#| fig-cap: "Observed vs. Predicted Over-Speeding: Model Accuracy Visualization"
#| echo: false

@@

[@fig-observed-vs-predicted] shows the relationship between predicted and observed over-speeding values. Predicted values align reasonably well with observed values at lower ranges (0–10 km/h). This indicates that the model performs adequately in capturing moderate over-speeding behaviours, where most of the data points fall close to the red dashed line representing perfect predictions.

@@

While the model provides reasonable predictions for average over-speeding behaviours, its performance diminishes at higher observed values, indicating potential limitations in its ability to capture outliers or extreme behaviours.

@@

### Wards with Lower Predictions
In contrast, Wards 5, 21, and 20 show substantially lower predicted over-speeding values, averaging below 3 km/h. These results suggest that predicted over-speeding behaviours are less prominent in these wards, potentially influenced by traffic conditions, enforcement presence, or other unobserved factors.

#### Implications of Predictions
While the predictions align with observed patterns in some wards, such as higher over-speeding in Ward 12, the variability across wards suggests that spatial characteristics may play a significant role. These predictions provide a baseline understanding of over-speeding behaviour at a ward level.

@@

## Ward-Level Disparities in Over-Speeding
The results show geographical disparities in over-speeding across Toronto school areas. For instance, Ward 12 emerges as a statistical outlier, with an average predicted over-speeding of 12.5 km/h. In  contrast, Ward 5 shows much lower predictions, averaging less than 3 km/h. This stark difference raises questions about factors such as road design, traffic enforcement, and even demographic variations. Ward 12 may be plagued by wide arterial roads inviting speed or does have outdated enforcement options. The low over-speeding in Ward 5 may hint at effective urban planning or stronger enforcement presence. It’s clear that “where” you drive in Toronto heavily influences “how” you drive. However, wards span cities in the 25 Ward model, and generalizing spaces of this size may also look over more specific area data.

Ward 4, which includes Parkside Drive—Toronto's most-ticketed location for speeding [@cbc2023], exhibited an intriguing pattern. Despite its reputation for dangerous speeding and expected dominance in overall speeding, the model indicated that over-speeding in Ward 4 was, on average, 4.77 km/h lower than the reference Ward 2. This discrepancy suggests the effectiveness of enforcement measures or specific traffic conditions in mitigating speeding behaviour.

@@

## Volume of Speeds
Higher traffic volumes seem to correlate with lower over-speeding rates. The consistently negative coefficient for volume across all models suggests that when more vehicles are on the road at any speed, the likelihood of speeding diminishes. One possible explanation is that congestion creates natural constraints, limiting a driver’s ability to exceed the speed limit alongside the speed of the drivers around them.

## Outliers 
The scatterplot of predicted vs. observed over-speeding values in [@fig-ward-predictions] reveals areas where the model struggles, particularly with extreme outliers. Predictions for high over-speeding values (e.g., above 40 km/h) are generally underrepresented, while predictions cluster heavily near lower values. This suggests the model may be oversimplifying the dynamics of extreme over-speeding events, which likely involve additional, unmeasured factors like driver behaviour or road conditions. Additionally, the residual errors in [@fig-residuals] highlight slight over-predictions for certain wards, hinting at potential biases in how ward-level effects were estimated.

@@

### Data and Method Limitations
The analysis does not account for date and time-specific patterns. Over-speeding frequently occurs during night hours or non-peak times when roads are less congested and school zone speeds only operate between 8 AM and 5 PM. Additionally, it is unclear whether the speed signs are specifically tracking school zone speeds or normal speeds. Some speed cameras included in the dataset were taken down partway through 2023, but they were retained in the analysis as long as they existed at any point during the year. While this decision did not significantly affect results, it did not clearly reflect temporal changes in camera presence.

The School Safety Zone Watch Your Speed Program sensors may occasionally track outliers such as ambulances or police vehicles, leading to exaggerated over-speeding measurements. Additionally, these systems do not log vehicle types, which could have provided more support into driving patterns. The analysis assumes that vehicles maintain a constant speed through the sensor area, which may not reflect real-world driving behaviour. Traffic patterns can be complex and consequently difficult to accurately model. The multi-linear regression model used could benefit from more layered or Bayesian models and uncertainty handling.

A major assumption in the modelling process is the uniform distribution of speeds within speed bins. However, this may not reflect real-world conditions, where driver behaviour often skews towards certain thresholds (e.g., staying just below enforcement limits). Similarly, this analysis aimed to examine trends in speeding around schools near traffic cameras, however it lacks in tracking speed demons and focuses on all speeds above the limit.

@@

## Next Steps and Recommendations

Future models could include weather data, as adverse conditions (e.g., rain, snow) are known to impact driving behaviour. Open Data Toronto provides weather datasets that could be merged with speeding metrics for more comprehensive analyses if joins are possible. Furthermore, wards with greater populations could have weight added to them according to size.

@@

Incorporating data on vehicle types could help identify whether certain vehicles (e.g., trucks, commercial vehicles) contribute disproportionately to over-speeding. Modified cars may tend to speed at greater velocity and more often than others due to adjusted, increased speed capacity, offering additional information on driver behaviour. Furthermore, there has been evidence that increases in vehicle height also correlate with speeding, as vehicle height may affect drivers' speed perception [@height].

As speeding trends can vary greatly over time, additional analysis spanning over several years can further support driver trends and behaviour. For example, speeding increased during the pandemic, likely due to the significant decrease in cars on the road, offering more leeway for speeding [@pandemic]. Having a stronger focus on speed demons may be of interest, as tracking driver behaviour of those who minimally speed may not be as telling of general speeder habits. Furthermore, considering the time of speed logging may improve and further detail the analysis, as speeding increases towards the peak hours and the night time [@nightSpeed] and it would better align with Toronto school zone speed laws from 8 am-5 pm.

@@

## Overview
This appendix outlines a survey methodology designed to improve the analysis of excessive speeding near Toronto school zones. With a $500K budget, this approach aims to gather data from drivers, pedestrians, and community members about speeding behaviours, awareness of enforcement measures, and perceived safety. By joining observational data with survey data, this methodology aims to improve analysis and informed policy recommendations.

@@

Vision Zero’s annual budget for public engagement and data-driven initiatives often ranges between 1-2 million, justifying a scaled-down 500K budget for this specific study. Based on these similar city-wide traffic safety initiatives the $500K is allocated to account for the following:

@@

- Thorough recruitment and incentivization strategies to ensure adequate participation.

@@

Drivers, particularly those aged 18-40 with an emphasis on young male drivers, are included to understand the behavior of frequent commuters in school zones. Pedestrians, such as parents, school staff, and local residents, provide analytics into the community's experiences and concerns. Finally, community advocates, including stakeholders involved in traffic safety initiatives, contribute broader perspectives on strategies to improve safety in high-risk areas.

### Behavioural Targeting
Beyond demographic data, the recruitment strategy will aim to identify drivers who:

- Exhibit risky behaviours, such as speeding or tailgating, based on self-reporting and survey data analysis.
